Appendix 3: Estimating species interactions

### 2015-11-05 Harris_2015/manuscript-materials/response-to-editor.md

Dear Mr. Harris,

Thank you very much for submitting your revised manuscript "Estimating species interactions from observational data with Markov networks" (Statistical Report) for review by Ecology. Many aspects of the presentation are improved. Unfortunately my assessment is that the new community dynamics simulations turned out to be a step in the wrong direction, making it harder rather than easier to assess the new method and compare it to older ones. I can see that this new material resulted from taking Reviewer #1's suggestion to heart, but the bottom line is that it didn't turn out well.

First, I don't think you needed to remove the extensive set of potentially very informative simulations that you had done previously but, if anything, add to them. (I know the 20-page Report limit is tight, but you have no limits in the supplement.) Evaluation of new methods when assumptions are not met builds naturally on first having evaluated the method when its assumptions are met, not skipping over that step. However, you have removed the lion's share of such comparisons, leaving only the simple 3 species case before diving into the community dynamics simulations. I was left unable to know how the method performs on its own terms before being faced with evaluation of how it performs on very complicated terms.

> I appreciate this perspective.  I felt very ambivalent about including these simulations in the first place, and had many of the same concerns that you raise.  I ultimately decided to include them because a substantial number of colleagues that read the preprint had similar concerns to Reviewer 1: they wanted to see how the different methods would perform on more realistic data sets and especially on data sets that were not simulated from a model that resembled a Markov network.  I do agree, however, that it is important to show that the model performs well on its own terms before diving into a more complicated model, and that omitting the simpler simulations made the paper worse. For this reason, I now include three different simulation regimes for the 20-species landscapes: 1. Simulating directly from a Markov network, 2. Simulating from a Markov network with spatial heterogeneity in the α parameters (sometimes called a conditional random field), and 3. Simulating with a dynamic model.

> For this revision, I considered a number of alternative simulation methods for the dynamic model that could address the concerns of Reviewer 1 and my colleagues, but they all had serious problems.  Using a set of GLMs as the generative model (as you suggest below) would either involve directed cycles (if β_ij≠β_ji) or would match the Markov network exactly (if β_ij=β_ji). Allowing directed cycles in the generative model would cause a number of problems. Most importantly, these models would not be identifiable from the presence-absence vectors they generated, as an infinite number of different parameter settings could all generate a given joint distribution. On the other hand, re-labeling the simulation network as a set of GLMs with equality constraints on the coefficients might have alleviated some readers' concerns, but only by hiding the fact that it was actually the same model.

Your proposed community dynamics model would raise a host of concerns. You state it is a large population approximation but then include demographic stochasticity, which is typically of diminishing importance for large populations and outweighed by environmental stochasticity, which you omit.

> I now refer to the differential equation as the model's deterministic backbone, rather than as an approximation.

The choice of parameters to draw from is arbitrary and ultimately discernible only by reading code, and the mutualism term appears to be arbitrary.

> This is true, but it was also true of all the other models I could envision where mutualism was guaranteed not to cause unbounded population explosions regardless of the parameters chosen.

You may disagree, and one could debate all aspects, but that is really the point: one would have wanted a simpler and more standard model so comparison of the statistical methods is not tangled in these issues.

> The simpler models are now included.

I can see you made a serious effort, but the outcome is not something I think readers will gain value from. The ultimate problem is that we are left with no comprehension of how the data generated by these processes suit the assumptions of the various methods. Does the model actually generate data that are well approximated by the Markov network assumptions or not? Trying to understand that would be a theoretical ecology exercise that, for your proposed model, does make sense to include in a paper about the statistical methods.

> Based on the number of readers that have asked for something along these lines (including Reviewer 1), I think that there is some value.  The new Figure 3 shows that the addition of population dynamics and per-capita interactions does weaken the relationship between the simulation coefficients and all of the estimation methods, but that it does not change the rank ordering of the different methods, which is an important result.

I suggest instead staying closer to the statistical models and including some additional cases to probe the method's robustness to violations of its assumptions and/or cases where one of the other parametric models in the competition generates the data. Which kinds of potential violations of assumptions to consider is for you to decide. For example, perhaps the cases of interest for the Markov network model would be if there really are 3-way interactions, or asymmetric interactions, or heterogeneity among sites in the model parameters, that contribute to the data generating process but are ignored in the analysis model. Indeed, describing what would constitute violations would help the reader. Another good way to probe would be to generate data from one of the other parametric models, such as the GLM.

> I now do this with the first two simulation types: the first type matches the model's assumptions and the second one adds heterogeneity among sites.  I now show that this violation decreases the models' R-squared values and increases their Type I error rates.

Note that often violations of assumptions do not create bias in parameter estimates (although sometimes they do) but rather incorrect assessments of uncertainty (inaccurate confidence intervals and Type I error rates in hypothesis tests). You have referred in multiple places to having applied a bootstrap as well as calculating Wald intervals, but it appears to me you have reported almost no results about validity of confidence intervals or hypothesis tests from those (there are a few results for the 3-species case). It would be important to do so.

> I now report approximate Type I error rates

It appears to me that you did not understand my point that you are presenting results that are integrated over parameter distributions, which appears still to be the case. What I meant is the following. Often one would want to see how a statistical method performs as a function of specific parameter values. For example, one might vary the value of an interaction coefficient (beta) from negative (competition) to zero to positive (mutualism) and run a set of simulations for each value of beta. This would allow evaluation of statistical power and accuracy of confidence intervals and Type I error rate as well as RMSE, all as a function of true parameter values. It would also allow assessment of whether the method works better for some parameter scenarios than others, which is not uncommon. Instead you are drawing parameters from distributions and then presenting performance assessments in aggregate from those results, such as the average R-squared over all cases of a given network size. Generally, and clearly in your case, such metrics can be written as expectations (integrations) over the distribution of parameters you sampled from. You are right there is nothing Bayesian about it, but I pointed out a Bayesian may feel justified in doing it because many Bayesian results take the same format. It is a question of presenting aggregated versus disaggregated results. Unfortunately, presenting only aggregated results leaves the reader unable to dig deeper into the disaggregated results. The analogy to sampling from an archipelago is lovely but irrelevant. It appears to me that since you have all the simulations at hand, and since you can place more detailed results in the supplement, you should do so. I leave it to you to decide how to do that.

> Showing how some of the results change with the value of beta is a great idea, and I have added several analyses along these lines in Figure 4 and Appendix 5.  In particular, I show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta, in addition to global estimates of R-squared and Type I error rates.  Unfortunately, I am dealing with a 210-dimensional system (20 alphas and 190 betas), which limits my options.  Holding 209 parameters constant at arbitrary values doesn't seem like the right answer, since the results would depend on which values were chosen.  Apart from the trivial case where all the coefficients except one are fixed at zero (in which case the indirect interactions that make this paper's methods necessary would not exist), there doesn't seem to be any particular line through the 210-dimensional space that deserves this kind of special attention. Likewise, there isn't a good way for completely free-form exploration of this large space.  For this reason, the best compromise in this revision seemed like marginalizing over all but one variable at a time for these analyses.

Some other minor notes

- I don't think you have stated that the betas are symmetric. It appears that beta_ij = beta_ji and hence each could be interpreted as half of the interaction strength. E.g. on line 80, it would be e^4 if I take the preceding model equation as correct and include both beta_ij and beta_ji. Please fix it up if that is not right.

> I've made this issue much clearer in the revision: see lines [[XXXX]].  It should now be unambiguous that the sum in the likelihood should only include each pair once, and that each pair must share a single value for $\beta_{ij}$.

- Both times that I have seen the title anew, the phrase "estimating species interactions from observational data" made me think it would be about time-series method, since that is a common kind of "observational data" from which people try to estimate interactions. I recommend adding a descriptive term like "species composition data" or "observational occurrence data" or "presence/absence data." You decide. This could enter the first Abstract sentence as well.

> This is another good point.  I've changed the title to focus on co-occurrence data.  Now that I've included inferential statistics in the paper and evaluated their frequentist properties, I've also changed the word "estimating" back to "inferring".

- Line 44 is not stated well. You could replace "begin with the assumption" by "test the hypothesis" or insert "that the hypothesis of interest is that [all pairwaise interactions...]" or some such change.

- Line 69: The phrase "how groups of species can co-occur" seems imprecise or incorrect. Maybe "to determine species occurence models"?

- Lines 86-87 are not a good lead-in to the point of the rest of this paragraph. It sounds like you are about to talk about species interactions, again, but really the point of the paragraph is that normalizing the probabilities involves a difficult summation.

- Lines 72 & 101. I was willing to overlook the fact that you didn't define the vector y explicitly in terms of its elements on line 72 because perhaps that is really obvious. However on line 101 you then state maximum likelihood as if it would be done from a single y vector (and a single likelihood term) rather than a matrix of observations. So now I think you need more explicit notation to clarify the arrangements of your variables.

> The revised manuscript now clarifies that each $\vec{y}$ vector represents
the different species that could occur at a single site and that the full
co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape." This should
clarify the relationship betweeen the $\vec{y}$ vectors and the matrix as well
as the relationship between the single-site likelihood and the landscape-level
likelihood (independence among sites).

- Line 130-132. This statement is unclear and unconvincing. Are you saying that the remoteness of the upper range of the prior justifies that the prior has no influence? That would not be correct since it is an issue of prior weightings in the range of heavy posterior weight. The simplest way to show a prior has little influence is to try a couple of different ones.

- Lines 136-137. Bootstrapping is a method to estimate uncertainty of an estimation procedure. It is unclear what you mean by using it to validate if a procedure gives "stable" estimates.

> The bootstrapping results have now been removed and replaced with the confidence intervals.

- Line 154: This statement is not clear.

- Figure 2B: This is odd notation because, if I follow, you have chosen a null symbol (0 with a slash) to indicate the event y_i = 0 but have chosen the label y_i to indicate the event y_i = 1. I think you want P[00], P[10], etc. or P[y1=0, y2=0], P[y1=1, y2=0], etc, or something like that. The caption states "relative probability" where I think you want "probability."

> This is an excellent point. The figure has been fixed (along with the caption).

- Figures in general: would benefit from titles on each subfigure.

> Subfigures now have titles

It's too bad we are in this unusual situation that your effort to respond to a reviewer made the manuscript worse. I will let you make another revision to get back on course. What I need you to do are: (1) remove the community dynamics model, put back in extensive results from the model of interest, and take a new approach to probing the model more thoroughly by including some scenarios with clearly defined violations of assumptions and/or data generated from one of the alternative statistical models. You may decide what to put in the online material since I recognize you are trying to stick to the 20-page report limit. (2) Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation. (3) Provide some disaggregated results. (4) Show how the regression steps work in about a page or two of supplement text, not requiring code reading.

### 2015-11-05 Harris_2015/manuscript-materials/response-to-reviewers.md

\setlength{\parskip}{0pt}

## Responses to Editor's comments

* The reviewers and I appreciate the work you have accomplished. Based on the reviews, we are willing to consider a revised version for publication in the journal, assuming that you are able to modify the manuscript according to the recommendations.

> **Response:** Thank you.

1. I would like to see a more conventional presentation of a new statistical method: state what kind of data needs to be modeled, present the model and estimation method, and then present the simulations and/or examples. This would mean placing your "Simulations" section after the section that should be labeled "Estimation" (rather than "Inferring..."), so that it comes as an evaluation of the method you have just explained. Actually what you have stated is a maximum likelihood estimation method and you have not addressed inference, i.e. how to decide uncertainty or significance.

> **Response:** I've re-ordered the material as you suggested (model, then estimation method, then examples).  I have also made a clearer distinction between parameter estimation and hypothesis testing throughout the manuscript (e.g. lines 185-191 discuss point estimates and inferential statistics separately).

2. On page two please go more directly to the method. As written it sounds like the issue is somehow to adapt partial correlations to non-Gaussian data, and I think this is a rather indirect and confusing explanation. What you have are multivariate 0/1 data and what you are presenting is a joint probability model for such data. This would provide a more direct path toward explaining to your contribution.

> **Response:** I now treat the partial correlations as an approximation to the discrete network (lines 141-154), rather than treating the discrete network as a generalization of partial correlations and introduce the method as a way to model the probability of multivariate presence-absence data under different parameter sets (lines 66-111).

3. Most of lines 84-96 are redundant and could be removed. This would save you space (and the reviewers may or may not have realized you were hewing to the 20-page limit for Statistical Reports).

> **Response:** Removed.

4. Please tread lightly on physics analogies and language and where possible use the corresponding statistical language. For example, the "partition function" is a normalizing constant.

> **Response:** I now refer to the partition function as a normalizing constant (line 386) and no longer discuss the energy function.

5. I think you could use some better paragraph choices, but I am not sure since you have no paragraph indentations. Please do not do that again.

> **Response:** Indentations have been added.  It may be worth updating the Society's [page on "Preparing LaTeX manuscripts for ESA journals"](http://esapubs.org/esapubs/latexTIPsESA.pdf) (e.g. to suggest adding `\setlength{\parindent}{1cm}` at the beginning of the manuscript and `\setlength{\parindent}{0cm}` before the references and figure captions) so that authors know to do this in the future.

6. Provide distinct labels for the two correlation-based methods to carry through the paper (line 137).

> **Response:** The distinctions between the different methods (including the ones added in response to reviewer comments) should now be clear throughout the paper (e.g. lines 192-203).

7. 24 simulations for a scenario is not very much.

> **Response:** I now perform 50 simulations per landscape size (150 total).  The bootstrap resampling performed in Appendix 3 indicates that the estimates reported in the Results section are stable with this many replicates (e.g. the 95% bootstrap CI indicates that Markov networks performed between 14 and 18 percent better than the next best method, which is a narrow enough range to have some confidence in the results).

8. The interpretation of your more complicated simulation and estimation exercise is fairly convoluted. Normally one would like to see a set of scenarios with fixed parameter values, but you are integrating over arbitrary distributions of parameter values in a way that a Bayesian might appreciate but really makes it difficult to know if the relative performance of the methods depends on the parameters. You could provide more elaborate results in the appendix. Please attempt to disentangle some of the results to provide greater understanding.

> **Response:** The simulation and estimation procedure should now be much clearer.  It could be thought of as simulating 150 ecologists that each collect data from a different archipelago, and then asking whether their results are consistent with the underlying biology and with one another.  There is no integration over any distribution, nor is there anything particularly Bayesian about the approach.  The bootstrapping analysis in Appendix 3 in particular produces frequentist confidence intervals showing that the relative performance of the different methods is insensitive to changes in parameter values.

9. Your use of a prior distribution for alpha and beta is a serious concern. It appears to be highly informative towards the range of true parameters that you drew from. On the other hand, I am not sure what you did because you stated you used a logistic distribution, but the paper you cited discusses use of t distributions for parameters of logistic regression. A potential drawback of your method is the number of parameters to estimate, so use of an informed prior is troubling. You must give results without this prior or present a very solid justification for using it or come up with another approach. Given that the reviewers and I find your method potentially valuable, then even if it turns out that estimation is difficult without informative priors, that would be an important result for readers to see clearly.

> **Response:** I've removed the citation, since I agree it wasn't clear. That paper justified the *t* prior in part as an approximation to the logistic, but that wasn't clear from my previous submission.  The new discussion on lines 128-137 is more directly on-point.  I've also substantially weakened the regularizer substantially (well beyond what was done in the Faisal et al. paper pointed out by Reviewer 2) and clarified its likely influence on the results.

10. At lines 163-165, please explain why maximum likelihood estimates should give a model with "exactly the observed occurence frequencies and co-occurrence frequencies".

> **Response:** This is now clarified (lines 105-107).

11. Lines 174-175. Typically statisticians talk about the two steps of estimation followed by inference for a given method. In this sentence you talk about "fitting" (estimation), by which you mean your method, and then "inferring", by which you actually mean other methods for comparison. And you have not actually stated how you would do inference with the new method (presumably likelihood ratios or such). Please re-write.

> **Response:** This distinction should now be clearer throughout the manuscript.  The new manuscript now discusses inferential statistics in two places, using approximate confidence intervals derived from the Hessian matrix (lines 168-169) and the bootstrap distribution Appendix 4.

12. Lines 178-181 and 185-190 need explaining, with rationale and sufficient details for a reader to repeat the work. This can go in the appendix if you prefer.

> **Response:** This material is now explained in the first three Appendices, along with code that a reader could use to replicate the analyses exactly.  The regularizer discussed on lines 178-181 of the old text is now chosen automatically by the `corpcor` package (lines 146-148 of the new text). The description of the "Pairs" model's $Z$-score (lines 157-158) should now be clearer as well.

13. I believe I followed lines 194-202 but they will be confusing for most readers. Again you can use space in the Appendix to clarify.

> **Response:** The section on model evaluation has been rewritten, and should be much clearer (see lines 170-181).  Additionally, the first three Appendices include heavily-commented code for replicating the analysis.

14. In lines 207-211, it is unclear for each method whether you are reporting a percentage of "significant" results or simply the percentage of correct signs (positive or negative). This is confusing because on the one hand the randomization methods aim explicitly for significance, but on the other hand you haven't stated how you would do a hypothesis test for your method (although clearly you could). As a result I am not sure if you are comparing apples to oranges here.

> **Response:** I now discuss the sign of the point estimate separately from its statistical significance.  See lines 183-191.

15. Please consider carefully if "Markov networks" is the label you want to provide for these models. At first glance it seems a trivial or ironic usage of the concept, since by assumption all vertices (species) in the graph are connected, so the Markovian notion of depending only on graph neighbors carries no impact. I am not sure if there is a more precise or otherwise better label, but I ask you to give this some consideration.

> **Response:** I've considered this, and I can't find a better name from the literature.  "Markov random field" still has "Markov" in the name; the term "Ising model" is not helpful to non-physicists and often implies a grid-structured network; "Boltzmann machine" is not helpful outside of one area of machine learning and often implies the inclusion of latent random variables, which are absent here.  My reading of the literature indicates that the terms "Markov network" and "Markov random field" are routinely used to describe fully-connected graphs (e.g. "Learning Factor Graphs in Polynomial Time and Sample Complexity", by Pieter Abbeel, Daphne Koller, and Andrew Y. Ng, bottom of page 1746).

16. Section 1 of the Appendix needs explicit equations

> **Response:** This whole Appendix has been replaced.

17. Your simulations seems to have a curious feature: you are drawing from distributions conditioned on having all species present. It is conceivable that this could introduce bias to your results. The conditional distribution you are actually sampling from may have different correlations (and hence parameter sampling distributions) than the full distribution. Please address this concern.

> **Response:** I no longer condition on the number of species; some landscapes now have fewer than 20 species.

18. Both reviewers note that you have missed substantial relevant literature.

> **Response:** This literature is now addressed more directly (see responses to reviewers below).

## Responses to Reviewer 1's comments

* I very much liked the manuscript: it has a clear point to make. As somebody who has used covariance matrices to make inferences about interactions, I agree that their interpretation can be spurious, for reasons pointed out by the author.

> **Response:** Thank you.

* My first major critical comment is that - if I understood correctly - the data are generated by the same Markov networks that are fitted to the data as one of the competing approaches. Thus it is not at all surprising that the Markov network approach works best in recovering the correct parameter values - the comparison between the approaches is not fair. A more convincing demonstration that the Markov network approach is able to infer interactions (not just associations) would be to use a more mechanistic model to generate the data. For example, a time series model of interactive species where the interactions would happen preferably at the individual level (as they do in nature; predators eat prey, etc.). Or at least a population level model where the current abundances of the species influence their growth rates.

> **Response:** I now simulate my landscapes from a time series model where each species' birth and death rates depend on the abundances of the other species on the landscape (see lines 118-124 and Appendix 2).

* My second major comment is that the ms seems to ignore a large amount of recent work in community ecology that has been focused on estimating co-occurrence [examples of such paper include Pollock, L.J. et al. (2014) Understanding co-occurrence by modelling species simultaneously with a Joint Species Distribution Model (JSDM). Methods in Ecology and Evolution 5, 397-406. Clark, J.S. et al. (2014) More than the sum of the parts: forest climate response from Joint Species Distribution Models. Ecological Applications 24, 990-999. Ovaskainen, O., Hottola, J. and Siitonen, J. 2010. Modeling species co-occurrence by multivariate logistic regression generates new hypotheses on fungal interactions. Ecology 91, 2514-2521]. Thus I think the author should do a better job in describing the state of the art.

> **Response:** I now include a JSDM's correlation matrix as a competing estimator (lines 159-164, Figure 3) and reference Pollock et al. (2014) in a number of places (lines 33, 64, 162-163, and 263).

* My third major comment relates to the fact that all partial covariances can be derived simply by inverting the covariance matrix (which I think is worth mentioning in the paper, now I think it is mentioned only in the Appendix). As many of the earlier methods (e.g. listed above) provide the full posterior distribution of the covariance matrix, it seems highly feasible to link to (some of) those earlier methods by using them to estimate the posterior distribution of the covariance matrix, and then inverting the covariance matrix. That would yield the full posterior distribution of the partial covariances, without the technical problem mentioned in the ms. Such comparison would seem better connected to recent literature than the regularized ridge regression.

> **Response:** I now include partial correlations calculated from the JSDM as well, inverting 1000 samples from the posterior distribution over correlation matrices (lines 163-164, Figure 3).

* I further note that while partial covariances may not have been used much in community ecology, they have been used e.g. in gene and protein interaction networks - I think the author should make links to this literature.

> **Response:** This is now noted on lines 145-146.

## Responses to Reviewer 2's comments

* The problem of species interactions is an important one, and Markov networks do indeed present a very promising avenue of research. The technical competency of the work appears strong.

> **Response:** Thank you.

1) I'd really like to see a real data example in here. The simulations are fine at an introductory level. But I think adding a real example would help illustrate the true worth of the methodology. A real data application (which doesn't have to be large) could show how using Markov networks produces a result which tallies well with our current understanding of interactions between species - in much the same way as the hypothetical example is presented.

> **Response:** I agree that this would be an important step, however I do not have access to any co-occurrence data sets where the signs and strengths of the interactions between species had been independently estimated.  I also don't have enough space to fit such an analysis in the short format for the journal.  Fortunately, the Azaele et al. (2010) paper I cite already includes an application of the method to a real data set, which should make it less important to do so here.

2) I do think there are some key references missing. The following are but two examples of inferring species interaction networks from data: [Aderhold et a. 2012, *Ecological informatics* and Faisal et al. 2010, *Ecological Informatics*]. The reader needs to know how the current paper compares with the above, and other similar works.

> **Response:** The manuscript now cites Faisal et al. (2010) on lines 142 and 151, and makes comparisons two several of the methods proposed in that paper (regularized generalized linear models and gaussian models [in the form of partial correlations]), in addition to the Markov network, a null model, the correlation matrix, and two versions of a latent variable model (see responses to Reviewer 1 above).  I did not find the spaially-explicit methods in Aderhold et al. (2012) to be as relevant for the kinds of spatially-implicit model comparisons being made in the paper, although it is possible that I misunderstood something.

* Of course, it is also perhaps worth noting that Markov networks are common for models of spatial correlation in Ecology.

> **Response:** This is now noted on line 53.

3) The beta symbol in Fig 2A should be beta_12; this then helps connect the specific example in the figure with the formula at the top of page 5.

> **Response:** The symbol has been modified as suggested.

4) Lines 172-173; the value of 0.2 for the ridge parameter should be justified.

> **Response:** The paper no longer uses a pre-specified penalty. Instead, it relies on the `corpcor` package to determine the appropriate shrinkage for each matrix (line 148).

5) Some comments on sensitivity would be welcome - especially with regard to Figure 3. We really should see something more on the uncertainty (or stability) of the estimated networks.

> **Response:** The revised manuscript includes several new analyses to address this issue.  First, the uncertainty in Figure 3's results is estimated with bootstrap resampling of the 50 simulated landscapes in each size class (Appendix 3).  Second, the sensitivity to sampling error is estimated in Appendix 4, using bootstrap resampling of the sites in one landscape.  Third, the manuscript now mentions the approximate confidence intervals that can be produced using the Hessian matrix (lines 168-169).

### 2015-11-05 Harris_2015/manuscript-materials/draft.Rmd
add fakedata and updates

A Markov network defines the relative probability of observing a given vector of
species-level presences (1s) and absences (0s), $\vec{y}$, as

@@

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). As described in Appendix 2, I randomly drew
the “true” coefficient values for each replicate landscape
so that most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on the local environment.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  I generated the final set of landscapes by simulating
the population dynamics of 20 interacting species, as described in Appendix 3.

@@

\noindent
\textbf{Twenty species.}
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) outperformed those
that did not.  Of these, the Markov network consistently performed the best.
As anticipated by @lee_learning_2012, generalized linear models closely
approximated the Markov network estimates, especially when the data sets were
very large. As reviewed in @gotelli_empirical_2009, however, most analyses in
this field involve fewer than 50 sites, so model performance on the smaller
landscapes where the gap was larger is most important. As shown in Appendix
[[X]], the standard errors associated with the estimates in Figure 3 are small
(less than 0.01), so the differences among methods should not be attributed to
sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here as
its probability of rejecting the null hypothesis for a pair of species where
$|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for simulations that
matched the model's assumptions, versus 0.15 for simulations that included
environmental heterogeneity [[Appendix]].  For the null model, these rates were
0.31 and 0.50, respectively---far higher than the nominal 0.05 rate. Figure 4C
shows, across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the interaction
with 95% confidence.  The null model makes such errors nearly 8 times as often
as the Markov network, even though it only reject the null hypothesis twice as
often overall [[Appendix]].  The Markov network's errors were also more
concentrated around 0, so it never misclassified strong interactions like the
null model did.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing species.
The tree (top) tends not to co-occur with either of the two shrub species, as
indicated by the strongly negative coefficient linking them. The two shrub
species also compete with one another, but more weakly (circled coefficient).
**B.** In spite of the competitive interactions between the two shrub species,
their shared tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). **C.** Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a Markov
network leads to correct identification of the negative shrub-shrub interaction
(circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta_{12}$ coefficient is consistent with competition between the two species.
**B.** The coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
frequencies of all possible co-occurrence patterns between the two species of
interest, as calculated in the previous panel. **D.** Without competition (i.e.
with $\beta_{12}=0$, each species would occur more often.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations across
the three simulation types.

\noindent \textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the (negative) correlation coefficient,
after controlling for landscape size. **C.** For any given interaction strength,
the null model was much more likely to misclassify its sign with 95% confidence
than the Markov network was.

@@

>>>>

A Markov network defines the relative probability of observing a given vector of
species-level presences (1s) and absences (0s), $\vec{y}$ at a site, as

@@

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). As described in Appendix 2, I randomly drew
the “true” coefficient values for each replicate landscape
so that most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on the local environment.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  I generated the final set of landscapes with an
abundance-based model that included per-capita interaction rates instead of
per-species interaction rates, as described in Appendix 3.

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.  
Of these, the Markov network consistently performed the best. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates, especially when the data sets were
very large. As reviewed in @gotelli_empirical_2009, however, most analyses in
this field involve fewer than 50 sites, so model performance on the smaller
landscapes where the gap was larger is most important. As shown in Appendix
5, the standard errors associated with the estimates in Figure 3 are small
(less than 0.01), so the differences among methods should not be attributed to
sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for
simulations that matched the model's assumptions, versus 0.15 for simulations
that included environmental heterogeneity (Appendix 5).  Coverage was similar
for other values of $\beta$. For the null model, these rates were 0.31 and 0.50,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows, across a
range of true interaction strengths, the probability that the null model or the
Markov network will predict the wrong sign of the interaction with 95%
confidence.  The null model makes such errors nearly 8 times as often as the
Markov network, even though it only reject the null hypothesis twice as often
overall (Appendix 5).  The Markov network's errors were also more concentrated
around 0, so it never misclassified strong interactions like the null model did.

@@

@@

\setlength{\parindent}{0cm}
\setlength{\leftskip}{0cm}

**Figure 1: A.** A small network of three competing species.
The tree (top) tends not to co-occur with either of the two shrub species, as
indicated by the strongly negative coefficient linking them. The two shrub
species also compete with one another, but more weakly (circled coefficient).
**B.** In spite of the competitive interactions between the two shrub species,
their shared tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). **C.** Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a Markov
network leads to correct identification of the negative shrub-shrub interaction
(circled).

**Figure 2: A.** A small Markov network, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta_{12}$ coefficient is consistent with competition between the two species.
**B.** The coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
frequencies of all possible co-occurrence patterns between the two species of
interest, as calculated in the previous panel. **D.** Without competition (i.e.
with $\beta_{12}=0$, each species would occur more often.

**Figure 3:** Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations across
the three simulation types.

**Figure 4: A.** The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the (negative) correlation coefficient,
after controlling for landscape size. **C.** For any given interaction strength,
the null model was much more likely to misclassify its sign with 95% confidence
than the Markov network was.
